---
title: "Advanced Functionality and Writing Extensions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Functionality and Writing Extensions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(StratPal)
```

Here we go through how the presented workflows can be generalized and expanded upon.

## Different forward models

In the examples, we used age-depth models from a carbonate platform simulated with CarboCAT lite. This can easily be expanded to any other sedimentary forward model (be it marine or terrestrial, siliciclastic, mixed, or carbonate system).

For this, age-depth information needs to be extracted from the forward model. The vectors of elapsed model time vs. accumulated height can then be handed over to `tp_to_adm` to define age-depth model objects. These can then be used in the pipelines as in the examples.

If the forward model included erosion, you need to first account for time intervals where sediment is first deposited and later removed. For this, first define a sediment accumulation curve using `tp_to_sac`, and pass it to `sac_to_adm` to turn it into an age-depth model. You can then use the resulting age-depth model for your analysis pipeline.

## Empirical age-depth models

Of course you can use empirical age-depth models in your pipeline. You can create them from tie points using `tp_to_adm` just as in the case with forward models.

The workflow in `StratPal` does not work with `multiadm` objects as defined by `admtools`. These objects store multiple age-depth models to account for uncertainty. You can however reduce their complexity, e.g. via `admtools::median_adm`, and use the resulting age-depth model for the modeling pipeline.

## Other ecological gradients and niche definitions

In the examples we used water depth as ecological gradient. This was because water depth was stored by the forward model used. The approach for niche modeling used in `StratPal` works for all types of niches. Potentially relevant niches where information can be extracted from forward models are e.g. substrate consistency, temperature, water energy, and many others.

We used a simple niche definition based on a probability density function of a normal distribution. This can be expanded to arbitrary niche definitions along a gradient.

For niche modeling, two components are needed:

1.  A function `niche_def` . It takes as input values of the gradient, and returns numbers returns numbers between 0 and 1 as output (where 1 corresponds to the taxon being in its preferred niche, and 0 to it being completely out of its preferred niche). It must take vectorized input, and return a vector of equal length
2.  A function `gc` that describes how the gradient changes with time. It must take vectorized inputs, and return a vector of equal length.

As long as these conditions are met, niche preferences along any gradient can be modeled using `apply_niche_pref`.

## Different models of phenotypic evolution

Other types of phenotypic evolution in the time domain can easily be incorporated. They should stick to the following conventions to be seamlessly integrated:

-   They should take a vector of times as first argument

-   The implementation should be for a continuous time version of the mode of interest, as the time domain is generally sampled at irregular times (see Hohmann et al. 2024, methods section, for details)

-   They should return a list with two elements: One named `t` that is a duplicate of the first argument handed to the function. The second should be named `y` and should contain the simulated trait values.

-   The output list `l` should be assigned both the class "list" and "timelist" via the command `class(l) = c("timelist", "list")` , see `random_walk` as example

The last two points make sure plotting via `plot.timelist` and `plot.stratlist` works seamlessly.

## Different models of event type data

We used (constant and variable rate) Poisson point processes to simulate event type data. Any model of event-type data can be used (e.g. with temporal correlation), as long as it returns a vector with the timing of the events

## Taphonomy for event type data

Taphonomic effects can be incorporated using using the `apply_taphonomy` function. This is based on the same principle as niche modeling, and makes use of the `thin` function, but was not shown in the examples.

## Transform different types of data

`strat_to_time` and `time_to_strat` are generic functions of the `admtools` package that can transform different types of data. Currently they transform phylogenetic trees (`phylo` objects), lists (`list` class), and numeric data such as the event type data (class `numeric`).

In general, any type temporal (resp. stratigraphic) data can be transformed. For this, you need define an `S3` class for the object you would like to transform, and then implement the transformation of this object for `time_to_strat` and `strat_to_time`.

If you would like to add transformations for new S3 classes to the `admtools` package, please see the contribution guidelines of the `admtools` package for details on how to contribute code.

## References

\* Niklas Hohmann, Joël R Koelewijn, Peter Burgess, Emilia Jarochowska. Identification of the mode of evolution in incomplete carbonate successions.bioRxiv 2023.12.18.572098; doi: [10.1101/2023.12.18.572098]([https://doi.org/10.1101/2023.12.18.572098),](https://doi.org/10.1101/2023.12.18.572098),) Data published under a [CC-BY 4.0]([https://creativecommons.org/licenses/by/4.0/)](https://creativecommons.org/licenses/by/4.0/)) license.
