---
title: "Integration with the FossilSim and paleotree packages"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Integration with the FossilSim and paleotree packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(StratPal)
library(admtools)
```

This vignette describes how to integrate the `StratPal` package with the `FossilSim` and the `paleotree` packages.

## Integration with the `FossilSim` package

As is, the `StratPal` package handles `fossils` objects as constructed by the `FossilSim` package. These objects represent fossils sampled along a tree (phylogeny). They can be transformed using the functions `apply_taphonomy` and `apply_niche`. This will remove fossil occurrences according to preservation potential and change on taphonomic conditions (or niche definition and change in gradient). Note that this corresponds to an nonparametric expansion of the variable recovery model as implemented in `FossilSim`.

As example, we examine the preservation of a phylogenetic tree with taxonomy and sampled fossils 12 km from shore in a carbonate platform.

### Stratigraphic basis

```{R}
# construct age-depth model
adm_12km = tp_to_adm(t = scenarioA$t_myr,   # 12 km from shore
                h = scenarioA$h_m[,"12km"],
                T_unit = "Myr",
                L_unit = "m")
plot(adm_12km, lty_destr = 0,
     main = "Age-depth model 12 km from shore",
     lwd_destr = 3)
T_axis_lab()
L_axis_lab()

# water depth 12 km from shore
t = scenarioA$t_myr
wd = scenarioA$wd_m[,"12km"]
gc = approxfun(t, wd)
plot(t, gc(t),
     xlab = "Time [Myr]",
     ylab = "Water depth [m]",
     type = "l",
     lwd = 3)
```


### Trees with fossils

For the biological components, we simulate a tree with taxonomy and fossils in the time domain using `FossilSim`:

```{R}
set.seed(42)
# simulate phylogenetic tree
tree = ape::rbdtree(birth = 3, death = 1, Tmax = max_time(adm_12km))
# simulate taxonomy along the tree
s = FossilSim::sim.taxonomy(tree = tree)
# simulate fossils based on taxonomy
sampling_rate = 40 # high sampling rate
f = FossilSim::sim.fossils.poisson(rate = sampling_rate, taxonomy = s)
# plot tree with taxonomy and fossil times
FossilSim:::plot.fossils(f, tree = tree, taxonomy = s, show.taxonomy = TRUE, rho = 0)
```

### Ecological effects

Lets assume all taxa have identical niches and a preferred water depth of around 60 m, and are rather tolerant to fluctuations in water depth. Then the tree with fossil samples looks like this
```{R}
my_niche = snd_niche(opt = 60, tol = 40, cutoff_val = 0)

f_niche = f |>
  rev_dir(ref = max_time(adm_12km)) |>
  apply_niche(niche_def = my_niche, gc = gc) |>
  rev_dir(ref = max_time(adm_12km))
# plot tree after accounting for taxon niches
FossilSim:::plot.fossils(f_niche, tree, taxonomy = s, show.taxonomy = TRUE, rho = 0)
```

Note that although there are many more lineages present at the end of the simulation, there are fewer fossils recovered. This is because the water depth is too low (the platform is drowning) to actually observe the taxa. Note that the `apply_niche` function is only applied to the `fossils` object and not to the tree and the taxonomy object, as they are not influenced by niche preferences.

### Stratigraphic effects

Now let us examine how the tree would be observed in the stratigraphic domain:

```{R}
## transform tree, fossils, and taxonomy into stratigraphyc domain
# using the age-depth model
tree_strat = time_to_strat(tree, adm_12km) # no transformation of time to age required
s_strat = s |>  # taxonomy object in the time domain
  rev_dir(ref = max_time(adm_12km)) |> # convert age to time
  time_to_strat( adm_12km , destructive = FALSE) |> # transform using age-depth model
  rev_dir(ref = max_height(adm_12km)) # transform back into age
f_strat =  f |> # same here
  rev_dir(ref = max_time(adm_12km )) |>
  time_to_strat( adm_12km , destructive = TRUE)|> # destroy fossils coinciding with gaps
  rev_dir(ref = max_height(adm_12km ))

FossilSim:::plot.fossils(f_strat, tree = tree_strat, taxonomy = s_strat, show.taxonomy = TRUE, rho = 0)
```

Note that this is now a tree in the stratigraphic domain, so the x axis label should correctly be called "Height below outcrop top". The accumulation of fossils on the right side is due to the stratigraphic condensation observable at 12 m height in the age-depth model.

### Combined stratigraphic and ecological effects

Naturally, both ecological and stratigraphic effects can be combined by applying the niche model to the `fossils` object: 

```{R}
## transform tree, fossils, and taxonomy into stratigraphyc domain
# using the age-depth model
tree_strat = time_to_strat(tree, adm_12km) # no transformation of time to age required
s_strat = s |>  # taxonomy object in the time domain
  rev_dir(ref = max_time(adm_12km)) |> # convert age to time
  time_to_strat( adm_12km , destructive = FALSE) |> # transform using age-depth model
  rev_dir(ref = max_height(adm_12km)) # transform back into age
f_strat =  f |> # same here
  rev_dir(ref = max_time(adm_12km )) |>
  apply_niche(niche_def = my_niche, gc = gc) |> # apply niche model 
  time_to_strat( adm_12km , destructive = TRUE)|> # destroy fossils coinciding with gaps
  rev_dir(ref = max_height(adm_12km ))

FossilSim:::plot.fossils(f_strat, tree = tree_strat, taxonomy = s_strat, show.taxonomy = TRUE, rho = 0)
```

### Technical notes

`FossilSim` uses age relative to the present, whereas `admtools` uses time. The function `admtools::rev_dir` accounts for this by reversing time to age and the other way around (relative to a reference point). Use this function on all object produced by `FossilSim` before entering and after leaving the `admtools` and `StratPal` ecosystem.

By default, axis labels of `FossilSim` are "Time before the present". In the context of `StratPal` and `admtools` this is misleading, as the present might refer to the end of the observation/the highest stratigraphic point. As of now, no option to override the default axis labels are available.

## Integration with `paleotree`

To combine the `StratPal` package with the `paleotree` package, please convert the `paleotree` format into the `FossilSim` format as described in the `paleotree` vignette of the `FossilSim` vignette:

```{r eval=FALSE}
vignette("paleotree", package = "FossilSim")
```

After conversion you can proceed as described in the section *Integration with the `FossilSim` package*

## References

*  Barido-Sottani J, Pett W, O'Reilly JE, Warnock RCM. Fossilsim: An r package for simulating fossil occurrence data under mechanistic models of preservation and recovery. Methods Ecol Evol. 2019; 10: 835–840. <https://doi.org/10.1111/2041-210X.13170>
*  Bapst, D.W. (2012), paleotree: an R package for paleontological and phylogenetic analyses of evolution. Methods in Ecology and Evolution, 3: 803-807. <https://doi.org/10.1111/j.2041-210X.2012.00223.x>
*  Warnock R, Barido-Sottani J, Pett W, Joseph O, Stolz U (2024). _FossilSim: Simulation and Plots for Fossil and Taxonomy Data_. R package version 2.4.1,
  <https://CRAN.R-project.org/package=FossilSim>
