---
title: "Introduction to the StratPal package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to the StratPal package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
library(StratPal)
library(admtools)
```

## Overview

Welcome to the `StratPal` package. This vignette provides an overview of the structure of the package and preliminaries needed to efficiently use it. We go through [*installation*](#installation), [*dependencies*](#dependencies), provide an overview of the available [*example data*](#example-data), [*piping*](#piping), and working with [*age-depth models*](#adm).

If you want to skip the introduction, you can also [get started](#getting-started) right away.

Throughout the vignettes there are several **tasks** that you can solve. They are not required for understanding the functionality of the package. Their aim is to explore the available models and data and develop an intuition for stratigraphic paleobiology.

## Installation {#installation}

### CRAN

To install the `StratPal` package from [CRAN](https://cran.r-project.org/), run

```{r, eval=FALSE}
install.packages("StratPal")
```

in the R console.

### GitHub

To install the `StratPal` package from GitHub, first install the `remotes` package by running

```{r, eval=FALSE}
install.packages("remotes")
```

in the R console. Then, run

```{r, eval=FALSE}
remotes::install_github(repo = "MindTheGap-ERC/StratPal",
                        build_vignettes = TRUE,
                        ref = "HEAD",
                        dependencies = TRUE)
```

to install the latest stable version of the package and all its dependencies.

## Dependencies {#dependencies}

The `StratPal` heavily relies on the age-depth modeling tools provided by the `admtools` package, which is *automatically* installed when you install `StratPal`.

To use its functionality, you need to run

```{r, eval=FALSE}
library(admtools)
```

before running any of the examples. Specifically, this is needed for plotting of age-depth models and trait evolution. Below we provide a brief overview of the functionality of the `admtools` package we need. If you want more information, you can browse through the package vignettes using

```{r, eval=FALSE}
browseVignettes(package = "admtools")
```

or by visiting the package website at <https://mindthegap-erc.github.io/admtools/>.

## Example data {#example-data}

`StratPal` comes with synthetic example data, which contains model parameters and outputs of a carbonate platform simulated using CarboCAT Lite ([Burgess (2013)](#References), [Burgess (2023)](#References)). The data is stored in the variable `scenarioA`, which is automatically available when the package is loaded. The structure of the data is described under `?scenarioA`.

The data is taken from scenario A in [Hohmann et al. (2023)](#References), the modeling procedure is described in detail in [Hohmann et al. 2024](#References), see therein for a chronostratigraphic diagram and a transect through the carbonate platform.

The data contains information on the eustatic sea level curve used for the model run, elapsed model time, as well information on accumulated sediment thickness, water depth, and bed thicknesses and facies at locations 2, 4, 6, 8, 10, and 12 km from shore in the simulated carbonate platform.

As an example, here is the eustatic sea level curve used for the model run:

```{r, fig.alt="Plot of a sinusoidal sea level curve."}
plot(x = scenarioA$t_myr,
     y = scenarioA$sl_m,
     type = "l",
     xlab = "Time [Myr]",
     ylab = "Eustatic sea level [m]",
     main = "Sea level curve used as model input")
```

## Piping {#piping}

In the vignettes of the `StratPal` package, we use the base R *pipe* operator `|>`. While this is not required to run the package, it simplifies the code and makes the underlying logic of a modeling *pipeline* clearer.

This functionality is available from R version 4.2 on.

### Motivation and usage

Consider the following code for simulating and plotting a random walk using the `random_walk` function:

```{r, fig.alt="plot of a random walk"}
set.seed(42)             # set seed for computational reproducibility
t = seq(0, 1, by = 0.01) # times where we evaluate the random walk
l = random_walk(t)       # simulate the random walk
plot(l, type = "l")      # line plot of the results
```

This code does the job, but it has some flaws: We introduced a lot of intermediate variables, which makes it hard to trace the logic of what we are trying to achieve: plotting a random walk. Using the pipe operator `|>` we can clarify the logic and simplify the code:

```{r, fig.alt="plot of a random walk, identical to the one above"}
set.seed(42)            # set seed for computational reproducibility
seq(0, 1, by = 0.01) |> # define times of simulation  
  random_walk() |>      # simulate random walk
  plot(type = "l")      # plot
```

You see that the code does the same thing: it plots a random walk, but it does so in one step without intermediate variables by chaining together the commands using the pipe `|>`. This becomes a powerful tool once we combine more and more components into longer modeling pipelines. It also makes the code readable, as you can simply read it from left to right, without having to track any intermediate steps.

Semantically, you can read the `|>` as "take the data on the left of `|>` and use it as the first argument in the function to the right of `|>`.

### Advanced usage

You can also use `|>` to pass arguments that are not in the first place. For this, simply replace the argument with a underscore `_`:

```{r}
# calculate deciles of normal distribution
seq(0, 1, by = 0.1) |>
  quantile(x = runif(100), p = _) # pass left hand side to the p argument
```

## Age-depth models {#adm}

The `StratPal` package use of age-depth modeling functionality from the `admtools` package. Here we go through some basics of dealing with age-depth models. For more details on available functionality you can browse through the package vignettes using

```{r, eval=FALSE}
browseVignettes(package = "admtools")
```

or visit the package website at <https://mindthegap-erc.github.io/admtools/>.

To get started, first load the package using

```{r}
library("admtools")
```

The `StratPal` package comes with some example data for age-depth models stored in the `scenarioA` variable, see the section on [example data](#example-data) or `?scenarioA` for details.

### Defining age-depth models

Let's start with defining the age-depth model 2 km from shore in scenario A. This can be done with `tp_to_adm` (tie points to age-depth model):

```{r}
t = scenarioA$t_myr       # extract time tie points
h = scenarioA$h_m[,"2km"] # get height tie points 2 km offshore in scenario A

# define age-depth model
# h[i] is the stratigraphic position at time t[i]
adm = tp_to_adm(t = t,          # tie points in time
                h = h,          # tie points at height
                T_unit = "Myr", # add time unit 
                L_unit = "m")   # add length unit
```

The nice thing about constructing age-depth models this way is that there is plenty of functionality available to extract data from age-depth models, plot them, or transform data using them.

### Plotting

Now you can plot the age-depth model using the basic `plot` command:

```{r, fig.alt = "plot of an  age-depth model, showing how the time and depth domain are connected"}
# plot age-depth model, see ?plot.adm for details
plot(adm,
     lwd_acc = 2,   # plot thicker lines for intervals with sediment accumulation (lwd = line width)
     lty_destr = 0) # don't plot destructive intervals/gaps (lty = line type)
T_axis_lab()        # add time axis label
L_axis_lab()        # add length axis label
title("Age-depth model 2 km from shore")
```

### Extracting information

There is a lot of functionality available to extract information from an age-depth model:

```{r}
get_total_duration(adm) # time interval covered by adm
get_total_thickness(adm) # sediment accumulated 
get_completeness(adm) # stratigraphic completeness
summary(adm) # some summary statistics
```

We can now use the pipe operator to do some first analysis of the age-depth model

```{r, fig.alt="histogram of hiatus durations 2 km offshore in scenario A"}

# plot histogram of hiatus durations
adm |>    
  get_hiat_duration() |>
  hist( xlab = paste("Hiatus duration", "[", get_T_unit(adm),"]"))
```

You can see that there are a 8 shorter hiatuses (below 100 kyr) and 2 long hiatuses with a duration of more than 500 kyr. To get a detailed list with information on hiatuses you can use `get_hiat_list`.

### Transforming data

Given a stratigraphic position, an age-depth model can tell us how old that positions is. Conversely, if we know the timing of an event, an age-depth model can tell us at what stratigraphic position that event will occur. This can be used to transform all types of data from the time domain to the stratigraphic domain and vice versa.

In `admtools`, the transformation of data is done by the functions `time_to_strat` (for transforming temporal data into stratigraphic data) and `strat_to_time` (for transforming stratigraphic data into temporal data). Details on how this is done and what types of data can be transformed can be found in the [vignette of the `admtools` package](https://mindthegap-erc.github.io/admtools/articles/admtools.html), for applications to stratigraphic paleobiology see the vignettes linked below under ["Getting started"](#getting-started).

## Getting started {#getting-started}

With the preliminaries out of the way, you can go to

```{r, eval=FALSE}
vignette("phenotypic_evolution")
```

for details on how to model stratigraphic paleobiology of phenotypic evolution, or explore the vignette online under [mindthegap-erc.github.io/StratPal/articles/phenotypic_evolution](https://mindthegap-erc.github.io/StratPal/articles/phenotypic_evolution.html).

Go to

```{r, eval=FALSE}
vignette("event_data")
```

for details on how to model stratigraphic paleobiology of event data such as individual fossils and first/last occurrences of taxa, or explore the vignette online under [mindthegap-erc.github.io/StratPal/articles/event_data](https://mindthegap-erc.github.io/StratPal/articles/event_data.html).

See also

```{r, eval=FALSE}
vignette("advanced_functionality")
```

for details on how to expand on the modeling pipelines described here, or explore the vignette online under [mindthegap-erc.github.io/StratPal/articles/advanced_functionality](https://mindthegap-erc.github.io/StratPal/articles/advanced_functionality.html).

## References {#References}

-   Burgess, Peter. 2013. "CarboCAT: A cellular automata model of heterogeneous carbonate strata." Computers & Geosciences. <https://doi.org/10.1016/j.cageo.2011.08.026>.

-   Burgess, Peter. 2023. "CarboCATLite v1.0.1." Zenodo. <https://doi.org/10.5281/zenodo.8402578>

-   Hohmann, Niklas; Koelewijn, Joël R.; Burgess, Peter; Jarochowska, Emilia. 2024. "Identification of the mode of evolution in incomplete carbonate successions." BMC Ecology and Evolution 24, 113. <https://doi.org/10.1186/s12862-024-02287-2>.

-   Hohmann, Niklas, Koelewijn, Joël R.; Burgess, Peter; Jarochowska, Emilia. 2023. "Identification of the Mode of Evolution in Incomplete Carbonate Successions - Supporting Data." Open Science Framework. <https://doi.org/10.17605/OSF.IO/ZBPWA>, published under the [CC-BY 4.0](https://creativecommons.org/licenses/by/4.0/) license.
