---
title: "Intoduction to the StratPal package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Intoduction to the StratPal package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(StratPal)
library(admtools)
```

## Overview

Welcome to the `StratPal` package. This vignette provides an overview of the structure of the package and preliminaries needed to efficiently use it. We go through _installation_, _dependencies_, provide an overview of the available _example data_, _piping_, and working with _age-depth models_.

If you want to skip the introduction, go to

```{r, eval=FALSE}
vignette("phenotypic_evolution")
```

for details on how to model stratigraphic paleobiology of phenotypic evolution. Go to

```{r, eval=FALSE}
vignette("event_data")
```

for details on how to model stratigraphic paleobiology of event data such as individual fossils and first/last occurrences of taxa.

## Installation

To install the `StratPal` package, first install the `remotes` package by running

```R
install.packages("remotes")
```

in the R console. Then, run

```R
remotes::install_github(repo = "MindTheGap-ERC/StratPal",
                        build_vignettes = TRUE,
                        ref = "HEAD",
                        dependencies = TRUE)
```

to install the latest stable version of the package and all its dependencies.

## Dependencies

The `StratPal` heavily relies on the age-depth modelling tools provided by the `admtools` package, which is _automatically_ installed when you install `StratPal`.

To use its functionality, you need to run

```{r}
library(admtools)
```


before running any of the provided examples. Below we provide a brief overview of the functionality of the `admtools` package we need. If you want more information, you can browse through the package vignettes using

```{r, eval=FALSE}
browseVignettes(package = "admtools") # opens in Browser
```

or by visiting the package website at https://mindthegap-erc.github.io/admtools/.

## Example data

## Piping

In the `StratPal` package and its vignettes, we make heavy use of the base R _pipe_ operator `|>`. While this is not required to run the package, it simplifies the code and makes the underlying logic of a modeling _pipeline_ clearer.

This functionality is available from R version 4.2 on.

### Motivation and usage

Consider the following code for simulating and plotting a random walk using the `random_walk` function:

```{r}
t = seq(0, 1, by = 0.01) # times where we evaluate the random walk
l = random_walk(t)       # simulate the random walk
plot(l, type = "l")      # line plot of the results
```

This code does the job, but it has some flaws: We introduced a lot of intermediate variables, which makes it hard to trace the logic of what we are doing: plotting a random walk. Using the pipe operator `|>` we can clarify the logic and simplify the code:

```{r}
seq(0, 1, by = 0.01) |> # define times of simulation  
  random_walk() |>      # simulate random walk
  plot(type = "l")      # plot
```

You see that the code does the same thing: it plots a random walk, but it does so in one step without intermediate variables by chaining together the commands using the pipe `|>`. This becomes a powerful tool once we combine more and more components into longer modeling pipelines. It also makes the code readable, as you can simply read it from left to right, without having to track any intermediate steps.
 
Semantically, you can read the `|>` as "take the data on the left of `|>` and use it as the first argument in the function to the right of `|>`.

### Advanced usage

You can also use `|>` to pass arguments that are not in the first place. For this, simply replace the argument with a underscore `_`:

```{r}
# calculate deciles of normal distribution
seq(0, 1, by = 0.1) |> quantile(x = runif(100), p = _) # pass left hand side to the p argument
```

## Age-depth models

The `StratPal` package makes heavy use of age-depth modeling functionality from the `admtools` package. Here we go through some basics of dealing with age-depth models. For more details on available functionality you can browse through the package vignettes using

```{r, eval=FALSE}
browseVignettes(package = "admtools") # opens in Browser
```

or visit the package website at https://mindthegap-erc.github.io/admtools/.

To get started, first load the package using

```{r}
library("admtools")
```

The `StratPal` package comes with some example data fro age-depth models, which are available in the variable `scenarioA`, see `?scenarioA` for details on it content.

### Defining age-depth models

Let's start with defining an age-depth model. This can be done with `tp_to_adm` (tie points to age-depth model):

```{r}
t = scenarioA$t_myr # extract time tie points
h = scenarioA$h_m[,"2km"] # get height tie points 2 km offshore in the carbonate plarform of scenario A from Hohmann et al. 2024

# define age-depth model
# h[i] is the stratigraphic position at time t[i]
adm = tp_to_adm(t = t, # tie points in time
                h = h, # tie points at height
                T_unit = "Myr", # time unit of adm
                L_unit = "m")   # Length unit
```


### Plotting

Now you can plot the adm using the basic `plot` command:

```{r}
# plot age-depth model, see ?plot.adm for details
plot(adm,
     lwd_acc = 2,   # plot thicker lines for accumulative intervals (lwd = line width)
     lty_destr = 0) # don't plot destructive intervals (lty = line type)
T_axis_lab() # add time axis label
L_axis_lab() # add length axis label
```

### Extracting information

There is a lot of functionality available to extract information from an age-depth model:

```{r}
get_total_duration(adm) # time interval covered by adm
get_total_thickness(adm) # sediment accumulated 
get_completeness(adm) # stratigraphic completeness
summary(adm) # some summary statistics
```
We can now use the pipe operator to do some first analysis of the age-depth model

```{r}

# plot histogram of hiatus durations
adm |>    
  get_hiat_duration() |>
  hist( xlab = paste("Hiatus duration", "[", get_T_unit(adm),"]"))
```

Here it becomes clear that there are a 8 shorter hiatuses (below 100 kyr) and 2 long hiatuses with a duration of more than 500 kyr.

### Transforming data

Given a stratigraphic position, an age-depth model can tell us how old that positions is. Conversely, if we know the timing of an event, an age-depth model can tell us at what stratigraphic position that event will occur. This can be used to transfrom all types of data from the time domain to the stratigraphic domain and vice versa.

In `admtools`, the transformation of data is done by the functions `time_to_strat` (for transforming temporal data into stratigraphic data) and `strat_to_time` (for transforming stratigrahic data into temporal data). Details on how this is done and what types of data can be transformed can be found in the other vignettes (see section below).

## Getting started

With the preliminaries out of the way, you can go to go to

```{r, eval=FALSE}
vignette("event_data")
```

for details on how to model stratigraphic paleobiology of events such as fossil occurrences and first/last occurrences. Go to 

```{r, eval=FALSE}
vignette("phenotypic_evolution")
```

for details on how to model stratigraphic paleobiology of trait evolution.
